---
title: "Applied Quantitative Methods Assignment 1 and 2"
author: "Caroline Oliver"
output:
  html_document: default
  pdf_document: default
---
## Basic R Questions
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Questions 1 to 4 - Basic Data File Questions

```{r}
# Set Up: download data file into RStudio
tallGrassPrairiePreserveData = read.csv(file = "http://dmcglinn.github.io/quant_methods/data/tgpp.csv", header = TRUE)
```
### Question 1: What are the names of the columns in the dataset?
```{r}
colnames(tallGrassPrairiePreserveData)
```

### Question 2: How many rows and columns does the data file have?
```{r}
# Number of Rows:
nrow(tallGrassPrairiePreserveData)

# Number of Columns:
ncol(tallGrassPrairiePreserveData)
```

### Question 3: What kind of object is each data column?
```{r}
sapply(tallGrassPrairiePreserveData, class)
```

### Question 4: What are the values of the the datafile for rows 1, 5, and 8 at columns 3, 7, and 10?
```{r}
# Row 1 Column 3:
tallGrassPrairiePreserveData[1,3]
# Row 1 Column 7:
tallGrassPrairiePreserveData[1,7]
# Row 1 Column 10:
tallGrassPrairiePreserveData[1,10]

# Row 5 Column 3:
tallGrassPrairiePreserveData[5,3]
# Row 5 Column 7:
tallGrassPrairiePreserveData[5,7]
# Row 5 Column 10:
tallGrassPrairiePreserveData[5,10]

# Row 8 Column 3:
tallGrassPrairiePreserveData[8,3]
# Row 8 Column 7:
tallGrassPrairiePreserveData[8,7]
# Row 8 Column 10:
tallGrassPrairiePreserveData[8,10]
```

## Plots - Questions 5 & 6

### Question 5: PDF of the relationship between the variables “scale” and “richness”
```{r}
plot(x = tallGrassPrairiePreserveData[,5], y = tallGrassPrairiePreserveData[,6], xlab = "Area of Quadrant (m^2)", ylab = "Richness", main = "Relationship between Scale and Richness", col = "red")
```

### Question 6: What happens to your plot when you set the plot argument log equal to ‘xy’?

```{r}
plot(x = tallGrassPrairiePreserveData[,5], y = tallGrassPrairiePreserveData[,6], xlab = "Area of Quadrant (m^2)", ylab = "Richness", main = "Relationship between Scale and Richness", col = "red", log='xy')
```
#### Adding "log = 'xy'" to the plot function changes the x axis (scale) back to log values which is how it is represented in the original data file. Also 4 values were ommitted from the log plot that were included in the non-logarithmic model in Question 5.


## Intermediate R Questions
### Question 1: what did the loops create?

#### The values stored in the output object are the averages of the three flower types sepal length and width, as well as, petal length and width arranged in matrix format.

### Question 2: Describe using pseudo-code how output was calculated:

#### Loop through 3 types of flowers represented as integers 1, 2, and 3 and set the variable iris_sp to have only the samples from one type of flower at a time. Then loop through the columns (features, i.e. sepal length and width & petal length and width) of the instantiated flower type from the first loop and assign x and y variables to a 0 value for summation purposes. If the number of rows in a flower types feature is greater than 0, loop through the rows of each feature for each flower type. The x variable is a sum of all values in the rows for a particular feature and y is the number of rows or values for that feature. Output is then calculated for each flower types certain feature by taking the quotient of x over y.

### Question 3: How can the objects output, x, and y could be renamed such that it is clearer what is occurring in the loop.

#### Other possible names:
#### output -> featureAverages
#### x -> featureSum
#### y -> numberOfFeatures

### Question 4: It is possible to accomplish the same task using fewer lines of code?

```{r}
data(iris)
sp_ids = unique(iris$Species)

output = matrix(0, nrow=length(sp_ids), ncol=ncol(iris)-1)
rownames(output) = sp_ids
colnames(output) = names(iris[ , -ncol(iris)])

for(i in seq_along(sp_ids)) {
  iris_sp = subset(iris, subset=Species == sp_ids[i], select=-Species)
  if (nrow(iris_sp) > 0) {
      for(j in 1:(ncol(iris_sp))) {
        x = sum(iris_sp[j])
        y = nrow(iris_sp)
        
        output[i, j] = x / y 
      }
    }
  }

output
```
#### For Question 4, I was able to decrease the number of loops by one by using the sum function to get the total of the values for a given flower's feature (x) and by using the nrow function to get the total number of data points for the given flower's feature (y). I then used those x and y values to calculate the average.

### Question 5: for loop that will produce a vector y that contains the sum of vector x up to that index of x
```{r}
x = 1:10
y = vector()
for(i in 1:length(x)){
  if(i==1)
  {
    y[i] = x[i]
  }
  else
    y[i] = y[i-1] + x[i]
}

# the updated y vector

y
```

### Question 6: Modify your for loop so that if the sum is greater than 10 the value of y is set to NA
```{r}
x = 1:10
y = vector()
for(i in 1:length(x)){
    if(i==1)
    {
      y[i] = x[i]
      
      if(y[i] > 10){
        y[i] = "NA"
      }
    }
    else if(y[i-1] <= 10){
      y[i] = y[i-1] + x[i]
      
      if(y[i] > 10){
        y[i] = "NA"
      }
    }
    else
      y[i] = "NA"
}

# updated y vector with intergers larger than 10 represented with NA
suppressWarnings(as.integer(y))
```

### Question 7: Place your for loop into a function that accepts as its argument any vector of arbitrary length and it will return y.
```{r}
incrementalVectorSummation = function(x){
    y = vector()
    for(i in 1:length(x)){
        if(i==1)
        {
          y[i] = x[i]
          
          if(y[i] > 10){
            y[i] = "NA"
          }
        }
        else if(y[i-1] <= 10){
          y[i] = y[i-1] + x[i]
          
          if(y[i] > 10){
            y[i] = "NA"
          }
        }
        else
          y[i] = "NA"
    }
    
    return(suppressWarnings(as.integer(y)))
}
```
#### Function call and output examples:
```{r}
incrementalVectorSummation(x = 1:10)
incrementalVectorSummation(x = 1:20)
incrementalVectorSummation(x = -5:10)
incrementalVectorSummation(x = 4:1)
```

#### *NOTE:*
##### The supressWarning function was used in Questions 6 and 7 because an error was being thrown about the NA values having to be coerced. This coersion did not effect the validity of the algorithms/functions output so I silenced the error messages for a cleaner output.
