---
title: "SpacialModels_HW5_CSO"
author: "Caroline Oliver"
date: "2/20/2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(vegan)
data(BCI)
data(BCI.env)
## UTM Coordinates (in metres)
BCI_xy = data.frame(x = rep(seq(625754, 626654, by=100), each=5), 
                    y = rep(seq(1011569,  1011969, by=100), len=50))
```
#### Split BCI data into rare and common datasets
```{r}
abu = colSums(BCI)
BCIrare = BCI[ , abu<=7]
BCIcommon = BCI[ , abu>80]
```

### Rare BCI
```{r}
BCIrareAbu = apply(BCIrare, 1, function(x) sum(x > 0))
hist(BCIrareAbu)

plot(BCI_xy, cex=BCIrareAbu/max(BCIrareAbu))

rare_species = BCIrare$Tetrathylacium.johansenii


col_brks = hist(BCIrareAbu, plot=F)$breaks
col_indices = as.numeric(cut(BCIrareAbu, col_brks))
cols = rev(terrain.colors(length(col_brks)))
plot(BCI_xy, cex=2, pch=19, col=cols[col_indices])

xy_dist = dist(BCI_xy)
max_dist = max(xy_dist) / 2
rare_dist = dist(rare_species)

plot(xy_dist, rare_dist)
abline(lm(rare_dist ~ xy_dist), lwd=3, col='red')
lines(lowess(xy_dist, rare_dist), lwd=3, col='pink')
abline(v = max_dist, col='red', lwd=3, lty=2)
```
#### The rare species I choose from the BCIrare Dataset is Tetrathylacium johansenii. As you can see from the solid red line in the graph above, there is evidence of over dispersal of this species. This is determined from the fact that the solid red line has a downward slope.

### Common BCI
```{r}
BCIcommonAbu = apply(BCIcommon, 2, function(x) sum(x > 0))
hist(BCIcommonAbu)

common_species = BCIcommon$Triplaris.cumingiana

plot(BCI_xy, cex=BCIcommonAbu/max(BCIcommonAbu))

col_brks = hist(BCIcommonAbu, plot=F)$breaks
col_indices = as.numeric(cut(BCIcommonAbu, col_brks))
cols = rev(terrain.colors(length(col_brks)))
plot(BCI_xy, cex=2, pch=19, col=cols[col_indices])

xy_dist = dist(BCI_xy)
max_dist = max(xy_dist) / 2
common_dist = dist(common_species)

plot(xy_dist, common_dist)
abline(lm(common_dist ~ xy_dist), lwd=3, col='red')
lines(lowess(xy_dist, common_dist), lwd=3, col='pink')
abline(v = max_dist, col='red', lwd=3, lty=2)
```
#### The common species I choose to single out from the BCIcommon Dataset is Triplaris cumingiana. As you can see from the solid, mostly horizontal red line in the graph above, there is little to no evidence of spacial dependence in this common species.


## Question 2
###Set Up
```{r}
library(gstat)
library(nlme)
sp_ids = c("Drypetes.standleyi", "Cordia.lasiocalyx", "Hirtella.triandra",
           "Picramnia.latifolia", "Quassia.amara",
           "Tabernaemontana.arborea", "Trattinnickia.aspera", 
           "Xylopia.macrantha")
```
### One Species Model
#### Base Model
```{r}
lm_oneSpecies = gls(Drypetes.standleyi ~ Cordia.lasiocalyx, data = BCI[ , sp_ids[1:2]])
plot(Variogram(lm_oneSpecies, form = ~ BCI_xy$x + BCI_xy$y))
```

#### Residuals for One Species Model
```{r}
resOne = residuals(lm_oneSpecies)
plot(dist(BCI_xy$x + BCI_xy$y), dist(resOne))
lines(lowess(dist(BCI_xy$x + BCI_xy$y), dist(resOne)), col='red', lwd=2)
abline(v = max_dist, col='red', lwd=3, lty=2)
```
#### Update of One Species Model using Nuggets
```{r}
oneSpecies_nug = update(lm_oneSpecies, corr=corExp(nugget=T))
oneSpecies_nug
plot(Variogram(oneSpecies_nug, form = ~ BCI_xy$x + BCI_xy$y, maxDist = max_dist))
```
### Coefficient Comparison
```{r}
lm_oneSpecies
oneSpecies_nug
```
#### The coefficients for both the intercept and the species are lower for the model that includes nuggets than the base model. A higher coefficient means that the predictor variable has a greater impact on the dependant variable so the model without the nuggets and with the greater coefficients would be the better model.

### ANOVA Comparison of One Species Models
```{r}
anova(lm_oneSpecies, oneSpecies_nug, test = F)
```
### Discussion
#### As you can see from the ANOVA results, the One Species GLS Base Model did better than the updated GLS Model using nuggets. Both the AIC and BIC is greater for the model that included the nuggets compared to the model without the nuggets which suggests that the base model is better since it has a lower AIC and BIC score.

### All Species Model
#### Base Model
```{r}
lm_allSpecies = gls(Drypetes.standleyi ~ Cordia.lasiocalyx + Hirtella.triandra + Picramnia.latifolia + Quassia.amara + Tabernaemontana.arborea + Trattinnickia.aspera + Xylopia.macrantha, data = BCI[ , sp_ids])
plot(Variogram(lm_allSpecies, form = ~ BCI_xy$x + BCI_xy$y))
```

#### Residuals for All Species Model
```{r}
resAll = residuals(lm_allSpecies)
plot(dist(BCI_xy$x + BCI_xy$y), dist(resAll))
lines(lowess(dist(BCI_xy$x + BCI_xy$y), dist(resAll)), col='red', lwd=2)
abline(v = max_dist, col='red', lwd=3, lty=2)
```

#### Update of All Species Model using Nuggets
```{r}
allSpecies_nug = update(lm_allSpecies, corr=corExp(nugget=T))
plot(Variogram(allSpecies_nug, form = ~ BCI_xy$x + BCI_xy$y, maxDist = max_dist))
```
### Coefficient Comparison
```{r}
lm_allSpecies
allSpecies_nug
```
#### The coefficients show to be equal between the two models for all variables included so the spacial error term did not have any impact what so ever on the coefficients.

### ANOVA Comparison of All Species Models
```{r}
anova(lm_allSpecies, allSpecies_nug, test = F)
```

### Discussion
#### From the ANOVA we can see that the GLS base model performed better than the updated model using nuggets. Similarly to the One Species Model compariosn AVOVA, both the AIC and BIC scores are lower for the base model suggesting that it will perform better than the model including nuggets.

### Adding of Spacial Data Discussion
#### For both models, one using only one species as a predictor and the second using a subset of species as predictors, the inclusion of spacial data did not seem to positively effect the outcome of the model. In both cases the ANOVA showed that the base model that did not incliude spacial data performed better. One reason for the spacial model performing worse could be that there was not a strong enough correlation between abundance of a species and its spacial location in the grid for it to be of any help in predicting the dependent species variable. Space/location may simply not play a role for the certain trees ability to grow/abundance.